# Exercise 8

The data in the table below were collected in order to assess whether a new vaccination was
successful in reducing the number of cases of influenza (flu).

These data result from a survey of 1000 people in the spring following the winter in question. Each
person was asked whether they had no vaccine, 1 flu vaccine injection or 2 flu vaccine injections.

* Let $V$ be a random variable equal to the number of flu injections that the person had.
* Let $F$ be a random variable indicating whether the person did have flu ($F = 1$) or not ($F = 0$).

## Contingency Table
```{r, echo = F}
data = matrix(c(24, 9, 13, 289, 100, 565), nrow = 2, byrow = TRUE)
rownames(data) <- c("F = 1", "F = 0")
colnames(data) <- c("V = 0", "V = 1", "V = 2")
xsq <- chisq.test(data)

con_table <- addmargins(data)
for (i in 1:2) {
  for (j in 1:3) {
    con_table[i,j] <- paste(data[i,j], paste0("(", xsq$expected[i,j], ")"))
  }
}
knitr::kable(con_table, caption = "Contingency Table with Expected Frequencies")
```

## Expected Frequencies 
**Estimate the expected frequencies under the assumption that $V$ and $F$ are independent.** \
If $P(F = 1 \mid V = 0) = P(F = 1 \mid V = 1) = P(F = 1 \mid V = 2) = P(F = 0)$, then we expect that

* $P(F = 1) \times 313$ people who didn't have flu without being vaccinated
* $P(F = 1) \times 109$ people who didn't have flu with $1$ flu vaccine injection
* $P(F = 1) \times 578$ people who didn't have flu with $2$ flu vaccine injection

$$\hat{P}(F = 1) = \frac{46}{1000} = 0.046$$
$$\hat{\mu}_{11} = \hat{P}(F = 1) \times 313 = 0.046 \times 313 = 14.398$$
$$\hat{\mu}_{12} = \hat{P}(F = 1) \times 109 = 0.046 \times 109 = 5.014$$
$$\hat{\mu}_{13} = \hat{P}(F = 1) \times 578 = 0.046 \times 578 = 26.588$$
$$\hat{\mu}_{21} = 313 - 14.398 = 298.602$$
$$\hat{\mu}_{22} = 109 - 5.014 = 103.986$$
$$\hat{\mu}_{23} = 578 - 26.588 = 551.412$$

## Pearson Residuals
```{r echo = F}
options(scipen = 1, digits = 2)
knitr::kable(xsq$residuals, caption = "Pearson Residuals")
```

$$r_{11}^{P} = \frac{24 - 14.398}{\sqrt{14.398}} = 2.53$$
$$r_{12}^{P} = \frac{9 - 5.014}{\sqrt{5.014}} = 1.78$$
$$r_{13}^{P} = \frac{13 - 26.588}{\sqrt{26.588}} = -2.64$$
$$r_{21}^{P} = \frac{289 - 298.602-}{\sqrt{298.602}} = -0.56$$
$$r_{22}^{P} = \frac{100 - 103.986}{\sqrt{103.986}} = -0.39$$
$$r_{23}^{P} = \frac{565 - 551.412}{\sqrt{551.412}} = 0.58$$

##  Standardized Pearson Residuals
```{r echo = F}
options(scipen = 1, digits = 3)
stdres <- xsq$residuals

for (i in 1:2) {
  for (j in 1:3) {
    stdres[i,j] = stdres[i,j] / sqrt((1-as.numeric(con_table[i,4])/1000)*(1-as.numeric(con_table[3,j])/1000))
  }
}  
  
knitr::kable(stdres, caption = "Standardized Pearson Residuals")
```

$$r_{11}^{S} = \frac{r_{11}^{P}} {\sqrt{(1-\hat{p}_{1+})(1-\hat{p}_{+1})}} = \frac{2.53}{\sqrt{(1-0.046)(1-0.313)}} = 3.13$$
$$r_{12}^{S} = \frac{r_{12}^{P}} {\sqrt{(1-\hat{p}_{1+})(1-\hat{p}_{+2})}} = \frac{1.78}{\sqrt{(1-0.046)(1-0.109)}} = 1.93$$
$$r_{13}^{S} = \frac{r_{13}^{P}} {\sqrt{(1-\hat{p}_{1+})(1-\hat{p}_{+3})}} = \frac{-2.64}{\sqrt{(1-0.046)(1-0.578)}} = -4.16$$
$$r_{21}^{S} = \frac{r_{21}^{P}} {\sqrt{(1-\hat{p}_{2+})(1-\hat{p}_{+1})}} = \frac{-0.56}{\sqrt{(1-0.954)(1-0.313)}} = -3.15$$
$$r_{22}^{S} = \frac{r_{22}^{P}} {\sqrt{(1-\hat{p}_{2+})(1-\hat{p}_{+2})}} = \frac{-0.39}{\sqrt{(1-0.954)(1-0.109)}} = -1.93$$
$$r_{23}^{S} = \frac{r_{23}^{P}} {\sqrt{(1-\hat{p}_{2+})(1-\hat{p}_{+3})}} = \frac{0.58}{\sqrt{(1-0.954)(1-0.578)}} = 4.16$$

## Association
**Does it appear that there is an association between the number of flu injections and catching the flu? Justify your answer.** \

**Method 1**: We can check the association by conducting a Chi-Square Test of Independence. \
Null hypothesis $H_0$: There is no association between the number of flu injections and catching the flu. \
If the null hypothesis $H_0$ is true, and all the $\hat{\mu}_{ij}$ are large enough ($\hat{\mu}_{ij} \geq 5$ for all i, j), then $X \mathrel{\dot\sim} \chi_{2}^{2}$.
$$\begin{aligned}
  \chi_{2}^{2} &= \Sigma_{i.j} \frac{(n_{ij} - \hat{\mu}_{ij})^2}{\hat{\mu}_{ij}} \\
  &= \frac{(24 - 14.4)^2}{14.40} + \frac{(9 - 5.01)^2}{5.01} + \frac{(13-26.59)^2}{26.59} + \frac{(289 - 298.60)^2}{298.60} + \frac{(100 - 103.99)^2}{103.99} + \frac{(565 - 551.41)^2}{551.41} \\
  &= 17.32
\end{aligned}$$

We can plot the Chi-Square Density function: \

```{r echo = F}
curve(dchisq(x, df = 2), from = 0, to = 20, 
      xlab = "Chi-Square", ylab = "Density", main = "Chi-Square Density (Degree of Freedom = 2)")
abline(v = 17.32, lty = 3)
```
 
By inspecting the plot, we find that $P(\chi_{2}^{2} \geq 17.32)$ is extremely small. We can use R to confirm that $P(\chi_{2}^{2} \geq 17.32) = 0.000173 \ll 0.005$ (using pchisq(17.32, 2, lower.tail = F)). Therefore, since the p-value is within the given significance level, we reject the null hypothesis.

On the other hand, we can conduct the Chi-Square Test using R directly:
```{r, comment = NA, echo = F}
chisq.test(data)
```
Since the p-value calculated by R is $0.000174 \ll 0.005$, which is small enough to reject the null hypothesis, we can prove that there is an association between the number of flu injections and catching flu. \

**Method 2**: We can test the association by estimating the conditional probabilities.
```{r echo = F}
options(scipen = 1, digits = 2)
library(knitr)
estimated_con_prob <- matrix(c(24/46, 289/954, 9/46, 100/954, 13/46, 565/954), nrow = 2)
rownames(estimated_con_prob) <- c("$\\hat{P}(V = i \\mid F = 1)$", "$\\hat{P}(V = i \\mid F = 0)$")
colnames(estimated_con_prob) <- c("$i = 0$", "$i = 1$", "$i = 2$")
kable(estimated_con_prob, escape = F, caption = "Estimated Conditional Probabilities of the Number of Doses of Vaccines Injected Given Catching the Flu or Not")

estimated_con_prob_2 <- matrix(c(24/313, 289/313, 9/109, 100/109, 13/578, 565/578), nrow = 2)
rownames(estimated_con_prob_2) <- c("$\\hat{P}(F = 1 \\mid V = i)$", "$\\hat{P}(F = 0 \\mid V = i)$")
colnames(estimated_con_prob_2) <- c("$i = 0$", "$i = 1$", "$i = 2$")
kable(estimated_con_prob_2, escape = F, caption = "Estimated Conditional Probabilities of Catching the Flu or Not Given Number of Doses of Vaccines Injected")
```

We can either compare the estimated conditional probabilities of the doses of vaccines injected given having flu or not presented in the table above (Table 4), or do it the other way around (Table 5). We can clearly see that a large proportion of people who caught flu did not inject the vaccine, while about $70\%$ of people who didn't catch flu at least injected one dose of vaccine. On the other hand, we can conclude that the proportion of people who injected one dose of vaccine catching flu is the same as the proportion of people who didn't inject vaccines at all catching flu. But the proportion of people catching flu among those who injected two doses of vaccines has decreased significantly (a quarter comparing to the former ones). Therefore, there is an obvious association between doses of vaccines injected and catching flu or not. Injecting both doses of vaccines may effectively reduce the risk of catching flu, while on the other hand, injecting merely one dose of vaccine may not necessarily be effective.

## Improvement
**Suppose that each person in this survey made their own decision whether to have $0$, $1$, or $2$ vaccine injections. Can you see a potential problem with this? What would be a better way to find out how effective is the vaccine at preventing flu?** \
The data collected may be unreliable if anyone among the sampled population reported the wrong data. It would further leads to the inaccuracy of our test result of independence / association of these variables.
\
A better way to conduct this research is that we randomly draw samples of the same size from people who haven't injected the vaccine, people who have had 1 injection of the vaccine, and people who have had 2 injections of the vaccine based on their medical record. Then, conduct a survey among these three sampled groups of people and see if they have had flu or not.